Tracing the cultural evolution of meaning at the semantics-pragmatics interface

### 2016-11-13 paper/paper-v2.tex
paper rewrite

An elegant formal approach to modeling the competition between expressivity and learnability is
the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
its simplest, discrete-time formulation, the RMD defines the frequencies $x'_i$ of each type
$i$ in a population at the next time step as a function of: (i) the frequencies $x_i$ of each
type $i$ before the update step, (ii) each the fitness $f_i$ of type $i$, and the probability
$Q_{ji}$ that an agent who wants to imitate, adopt or learn the type of an agent with type $j$
ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}
\end{align}
The replicator mutator dynamic consists of two components: fitness-based selection and
transmission biases. This becomes most transparent when we consider an equivalent formulation
in terms of a step-wise application of the discrete-time replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St} and subsequent multiplication with a mutation matrix
$Q$ of the initial population vector $vec{x}$:
\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i \ \ \text{, where} \\
  (\text{M}(\vec{x}))_i & = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i \ \  \text{and}  \\
        \left ( \text{RD}(\vec{x}) \right )_i 
        & = \frac{x_i f_i}{\sum_k x_k f_k}
\end{align}

thus conceived: replication then mutation; if $f_i$ is the same for all $i$ we get only
multiplication by $Q$, if $Q_{ji} = 1$ exactly when $i=j$, we get the RD, la di da

main question what are types, what is fitness, what is $Q$ answered next

@@

\subsection{Lexica and linguistic behavior}\label{sec:languages+use}
Lexica codify the truth-conditions of a language's expressions. Following \citet{franke+jaeger:2014}, a convenient way to represent lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs (meanings) and $M$ a set of messages (forms available in the language).

>>>>

An elegant formal approach to modeling the competition between expressivity and learnability is
the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
$i$ before the update step, (ii) the fitness $f_i$ of each type $i$, and (iii) the probability
$Q_{ji}$ that an agent who wants to imitate, adopt or learn the type of an agent with type $j$
ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
\end{align}
The replicator mutator dynamic consists of two components: fitness-based selection and
transmission biases. This becomes most transparent when we consider an equivalent formulation
in terms of a step-wise application of the discrete-time replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St} and subsequent multiplication with a mutation matrix
$Q$ of the initial population vector $\vec{x}$:
\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i \ \ \ \ \text{and} \ \ \ \ 
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\sum_k x_k f_k}\,.
 \end{align*}
 If the transmission matrix $Q$ is trivial in the sense that $Q_{ij}=1$ whenever $i=j$, the
 dynamic reduces to the replicator dynamic. The replicator dynamic is a model of fitness-based
 selection in which the relative frequency of type $i$ will increase with a gradient
 proportional to its average fitness in the population. The replicator dynamic is popular and
 versatile because it can be derived from many abstract processes of biological and cultural
 transmission and selection \citep[for overview and any derivations
 see][]{Sandholm2010:Population-Game}, including conditional imitation
 \citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
 learning \citep[e.g.][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If $f_i$
 is the same for all types $i$, the replicator step is the identity map
 $ \left ( \text{RD}(\vec{x}) \right )_i = x_i$ and the dynamic reduces to a process of
 iteration of the transmission bias encoded in $Q$. In this way, the process in
 (\ref{eq:RMD_discrete}), equivalently (\ref{eq:RMD_discrete_recast}), contains a model of
 iterated learning \citep{griffiths+kalish:2007}.

Where our goal is an application of this dynamic to the case of co-evolution of semantic
meaning and pragmatic use, we need to fix what the relevant types are, how fitness is measured
and how the mutation matrix is computed. These issues will be addressed, one by one, in the
following.

@@

\subsection{Types: Lexica and linguistic behavior}
\label{sec:languages+use}

Lexica codify the truth-conditions of a language's expressions. Following \citet{franke+jaeger:2014}, a convenient way to represent lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs (meanings) and $M$ a set of messages (forms available in the language).
